---
title: "Building blocks of R"
# subtitle: "Getting started"
author: "Mikhail Dozmorov"
institute: "Virginia Commonwealth University"
# date: "`r Sys.Date()`"
date: "08-13-2020"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css", "xaringan-my.css"]
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "midnightblue",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "500", "500i"),
  code_font_google   = google_font("Droid Mono"),
  link_color = "#8B1A1A", #firebrick4, "deepskyblue1"
  text_font_size = "28px"
)
```

## Variables

- **Scalars** (0-dimensional): `a = 42`, `b = a / 7`

- **Vectors** (1-dimensional): `b = c(12, 14, 16)`
    - Do not use `c` as a variable name! Check `?c` - 'concatenate' function
    - Access vector element as `b[2]` (returns 14)

- **Matrices** (2-dimensional):
```{r}
mtx = matrix(data = c(3, 1, 3, 2, 3, 2), ncol = 2)
mtx
```

---
## Data frames

- **Data frames**: tables or 2-dimensional arrays. Think matrices that can hold different data types
    - The column names should be non-empty
    - The row names should be unique
    - The data stored in a data frame can be of numeric, factor or character
```{r}
dat = data.frame(Column.1 = c(3, 1, 3), Column.2 = c("2", "3", "2"))
dat
```

- Address any element as `dat[3, 2]`. `[]` contain row/column indices. Equivalent to `dat[3, "Column.2"]`, `dat$Column.2[3]`
- Compare `class(dat$Column.1)` and `class(dat$Column.2)`

---
## Lists

- **Lists**: objects containing elements of different types
    - Each list element can be of different length
    
```{r}
lst = list(A = rep(2, 5), B = seq(1:10), C = letters)
lst
```
- Address any element as `lst[1]` (or, `lst["A"]`)
- Address _the content of any element_ as `lst[[1]]` (or, `lst[["A"]]`, `lts$A`)

---
## `R` basics - assignment

- We often need to save a function's result or output. For this we use the assignment operator: ` <- `, preferred over ` = `  
- For example, when you import a CSV file you need to give it a name: 
```{r, eval=FALSE}
scores <- read.csv(file="scores.csv")
``` 
 
Now we can use `scores` as an argument to other functions. For example, compute summary statistics for each column in the data:

```{r, eval=FALSE}
summary(scores)
```

.small[**Note**: Use `Alt + -` (Win) or `Option + -` (Mac) in RStudio to quickly insert ` <- `. You can also use `=` for assignment.]

---
## `R` basics - functions

- A function is a set of statements organized together to perform a specific task
    - **Name** - the actual name of the function, e.g., `summary()`, `mean()`
    - **Arguments** - values passed to the functions. Argument-less functions exist
    - **Code** - actual code of the function
    - **Return value** - the result of the function's code execution

```{r eval=FALSE}
read.csv(file="scores.csv")
```

`read.csv` is a function to import a CSV file and `file` is an argument that specifies which file to import

R has a large number of built-in functions, and the user can create their own functions

---
## `R` basics - running functions

- From the R console - type the function and hit Enter
    - One function at a time, not efficient

- Using an `R` script - a text file that contains all your `R` functions/code
    - `R` scripts allow you to save, edit, reproduce and share your code
    - R scripts stored in files with `.R` extension
    - Run the whole script as `source("script_name.R")`, or, from command line, `Rscript script_name.R`

---
## Basic plotting

```{r}
plot(rnorm(100), type="l", col="blue")
```

---
## Basic plotting

```{r}
hist(rnorm(1000))
```

---
## Basic plotting

```{r}
mtx <- data.frame(A = rnorm(100), B = runif(100, min = -2, max = 2))
boxplot(mtx)
```

---
## If-else statement

Conditional code execution

```{r}
# set.seed(1) # Use to initialize random number generator
random_guess <- sample(c(0, 1), size = 1) # Random 0 or 1
if (random_guess == 0) {
  print("I love cats!")
} else {
  print("I love dogs!")
}
```

---
## For loop

Repetitive code execution

```{r}
all_i <- c(1:5)
for (i in all_i) {
  cat(i)
}
```

`apply` family of functions allow for most efficient repetitive operations
    - `apply()`, `lapply()` , `sapply()`, `vapply()`, `mapply()`, `rapply()`, and `tapply()` functions

---
## `R` basics - packages

- All functions belong to *packages*. The `read.csv` function is in the `utils` package.

- `R` comes with about 30 packages (called "base `R`"), but as of August 2020 there are over 16,000 CRAN packages and over 1,900 Bioconductor packages

- Example: `ggplot2` is a popular package that adds functions for creating graphs in a different way than what base `R` provides

- To use functions in a package, the package must be installed and loaded. (They're free)
- You only _install_ a package once
- You _load_ a package whenever you want to use its functions

---
## Installing packages

- `install.packages` - installs packages from CRAN, e.g., `install.packages("BiocManager")`

- `remotes` package - installs R packages from GitHub, GitLab, Bitbucket, Bioconductor, or plain 'subversion' or 'git' repositories. E.g., `remotes::install_github("tidyverse/ggplot2")`

- `BiocManager::install()` - Install or update Bioconductor, CRAN, or GitHub packages

- RStudio point-and-click interface

---
## Loading packages

- `library()` will load the package, e.g., `library(readxl)` or `library("readxl")`
    - But, when installing packages, always use parentheses, e.g., `install.packages("readxl")`

- `require()` will load the package, and, if success, return TRUE. Useful in `if` statement, e.g.

```
if (require(ggplot2)) {
  install.packages("ggplot2")
}
```

---
## Getting help

- Get an overview of all functions in a package: `help(package = "dplyr")`
    - Bioconductor packages have vignettes, short tutorials on package-specific tasks. Browse them, e.g., `browseVignettes(package = "limma")`

- Use `?function_name` to get help on a function from a _loaded_ package. E.g., `?boxplot` (same as `help(boxplot)`)
    - Use `example(boxplot)` to see how the function can be used

- Use `??function_name` to search for the function across all installed packages, even not loaded. E.g., `??ggplotly`

- Search engine is your best friend on many things

---
## Useful ways of getting data in and out of R

- Base functions: `read.table`, `read.csv`, `write.table`, `write.csv`

- Tidyverse way, `readr` package: `read_table`, `read_csv`, `read_tsv`, `write_csv` ...

- For fixed-width files, use `read.fwf` or `readr::read_fwf` funcitons

- For reading/writing Excel files, use `readxl` and `writexl` packages, `read_xlsx`, `write_xlsx` functions
    - Remember that `.csv` is the preferred text-based format that opens in Excel

.small[https://readr.tidyverse.org/

https://readxl.tidyverse.org/

https://CRAN.R-project.org/package=writexl]


---
## R is more than a programming language

Numerous packages are available to extend R functionality

- Publication-quality figures, documents in Word, PDF, and HTML formats (Rmarkdown)
    - Templates for journal articles are available

- Presentations, from basic (`ioslides`, `beamer`) to advanced (`xaringan`)

- Web sites for blogs (`blogdown`), books (`bookdown`), packages (`pkgdown`)
    - Templates for CV, resume, thesis are available

- Dynamic web applications using Shiny

- Interface with other programming languages, like C (`Rcpp`), Python (`reticulate`)

- Many more cool usages...